Open archaeology, open source?

### 2023-06-27 analysis/_01-introduction.qmd
Properly formatted bibliographic references

In their seminal vision for open archaeology, Beck and Neylon (2012: 480-481) identified the movement as comprising a series of principles and practices "predicated on promoting open redistribution and access to the data, processes and syntheses generated within the archaeological domain" with the aim of "maximizing transparency, reuse and engagement while maintaining professional probity". Open archaeology therefore promotes more thoughtful scholarly communication practices, most notably in operations relating to publishing, data-sharing, education, and review processes. Archaeologists are also actively engaged in open source software development as means of sharing their research processes and creating tools and resources for general use.

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation, which publishes a widely accepted defitnion of "open" in the context of scholarly communication, explicitly refers to the definition of "open source" published by the Open Source Initiative, an authoritative open source advocacy group (@openknowledgefoundation2023; @opensourceinitiative2007). The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plaintext, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs signififcantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL, and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers (Kelty 2008:xx).

The relationship between open science and open source is also complicated by rhetorical claims that, based on our personal experiences, are discordant with how academics actually _do_ open source. Does academic open source actually make research processes more transparent and improve research outcomes? Is it actually boosting efficiency by establishing a common store of knowledge and productive code? Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass? Despite hopeful aspirations espoused by open archaeology advocates (Kansa  et al. 2014; Kintigh et al. 2015; others), these outcomes are not a given. We believe that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. However, practical circumstances and systemic value regimes that frame what it means to work as an archaeologist presently inhibit the potential for radical transformation, even among open science's most ardent supporters.

>>>>

In their seminal vision for open archaeology, Beck and Neylon [-@beck2012 480-481] identified the movement as comprising a series of principles and practices "predicated on promoting open redistribution and access to the data, processes and syntheses generated within the archaeological domain" with the aim of "maximizing transparency, reuse and engagement while maintaining professional probity". Open archaeology therefore promotes more thoughtful scholarly communication practices, most notably in operations relating to publishing, data-sharing, education, and review processes. Archaeologists are also actively engaged in open source software development as means of sharing their research processes and creating tools and resources for general use.

However, academic open source has a complicated relationship with open source as practiced by professional software developers, which has its own distinct history and is framed by different objectives, challenges, and value regimes. Despite this, the open science movement, within which open archaeology emerged, draws direct inspiration from open source. For instance, the Open Knowledge Foundation [-@openknowledgefoundation2023] publishes a widely accepted defitnion of "open" in the context of scholarly communication that explicitly refers to the definition of "open source" published by the Open Source Initiative [-@opensourceinitiative2007], an authoritative open source advocacy group. The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plaintext, line-resolution version control, emphasis on formal licencing, the general hacker aesthetic). However, academic work, including the development of academic software, differs signififcantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, openSSL, and the Firefox web browser. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional warrants and interests. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers [@kelty2008 xx].

The relationship between open science and open source is also complicated by rhetorical claims that, based on our personal experiences, are discordant with how academics actually _do_ open source. Does academic open source actually make research processes more transparent and improve research outcomes? Is it actually boosting efficiency by establishing a common store of knowledge and productive code? Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass? Despite hopeful aspirations espoused by open archaeology advocates [see @kansa2014; @kintigh2015], these outcomes are not a given. We believe that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. However, practical circumstances and systemic value regimes that frame what it means to work as an archaeologist presently inhibit the potential for radical transformation, even among open science's most ardent supporters.

### 2023-06-27 analysis/_02-background.qmd
Properly formatted bibliographic references

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development (@kelty2008: chapter 3, especially starting at page 99). By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision (Kelty 2008 on the emacs saga). An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles. A good test of whether a contribution should be included in a published software release is whether it is functional (Kelty 2008: 220). Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes (@coleman2012; O'Neil 2009: the debian saga).

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants (O'Neil 2009). They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience (@kelty2008; Ratto dissertation and derived works).

@@

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" (@openknowledgefoundation2023). This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates (including archaeologists, as elicited by Beck and Neylon 2012; Marwick et al. 2017) claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scienctific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

The idea that scientistis should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals (@mirowski2018). Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by digital telecommunications systems that host files, document processes, facilitate co-working opportunities, and perform automated processes. However, these systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science.

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain (@tennant2020). But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (inlcuding: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software.

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scienctific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards (cf. @kansa2014; @kintigh2015; @roosevelt2015) distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, @faniel2013 :299-301, @atici2013 :676-677, @huggett2018, @sobotkova2018, and @opitz2021 demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. Here we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.

>>>>

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development [@kelty2008: chapter 3, especially starting at page 99]. By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision [@kelty2008 on the emacs saga]. An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles. A good test of whether a contribution should be included in a published software release is whether it is functional [@kelty2008 220]. Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes [@coleman2012; @oneil2009 on the debian saga].

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants [@oneil2009]. They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience [@kelty2008, Ratto dissertation and derived works].

@@

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" [@openknowledgefoundation2023]. This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates (including archaeologists, as elicited by @beck2012 and @marwick2017) claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scienctific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

The idea that scientistis should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals [@mirowski2018]. Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by digital telecommunications systems that host files, document processes, facilitate co-working opportunities, and perform automated processes. However, these systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science.

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain [@tennant2020]. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (inlcuding: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed, papers have been published, and credit has been allocated, it is common for scientists to mark their projects as finished and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly in perpetuity, and contributors may remain actively or sporadically involved to satisfy users' needs, or to direct users to derivative and functional forks of abandoned software.

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scienctific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards [cf. @kansa2014; @kintigh2015; @roosevelt2015] distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, Faniel et al. [-@faniel2013 299-301], Atici et al. [-@atici2013 676-677], @huggett2018, @sobotkova2018, and @opitz2021 demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. Here we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.

### 2023-06-27 analysis/_03-data_methodology.qmd
Refining the intro and background some more.
Properly formatted bibliographic references

Our findings are based on an exploratory analysis of open-archaeo, a list of open source archaeological software and other digital resources. As of right now it includes 408 items, which are rendered for the web at [open-archaeo.info](open-archaeo.info). We compiled the dataset by browsing collaborative software development platforms such as GitHub, GitLab and Codeberg, and by tracing connections to other personal, professional, and institutional websites that describe and host additional archaeological software. This entailed manually crawling through users' profiles, particularly those who identify as archaeologists or who have contributed to tools that pertain to archaeological work. We supplemented this quasi-systematic collection strategy with word-of-mouth contributions made by interested individuals who identified relevant work that we initially overlooked.

@@

For the purpose of this study, we augmented the core open-archaeo dataset with data obtained from GitHub's public API. That is, for each project in open-archaeo with a GitHub repository, we added information on individual contributors, commits, issues, etc. from the API using the R package `gh` [@gh]. This does reduce our sample size slightly; about an eighth of projects don't use version control, at least not publicly, but amongst those that do, GitHub is by far the most popular host ([@tbl-repo-hosts]).

>>>>

Our study comprises an exploratory analysis of open-archaeo, a list of open source archaeological software and other digital resources. As of right now it includes 408 items, which are rendered for the web at [open-archaeo.info](open-archaeo.info). We compiled the dataset by browsing collaborative software development platforms such as GitHub, GitLab and Codeberg, and by tracing connections to other personal, professional, and institutional websites that describe and host additional archaeological software. This entailed manually crawling through users' profiles, particularly those who identify as archaeologists or who have contributed to tools that pertain to archaeological work. We supplemented this quasi-systematic collection strategy with word-of-mouth contributions made by interested individuals who identified relevant work that we initially overlooked.

@@

For the purpose of this study, we augmented the core open-archaeo dataset with data obtained from GitHub's public API. That is, for each project in open-archaeo with a GitHub repository, we added information on individual contributors, commits, issues, etc. from the API using the R package `gh` [@csardi2023]. This does reduce our sample size slightly; about an eighth of projects don't use version control, at least not publicly, but amongst those that do, GitHub is by far the most popular host ([@tbl-repo-hosts]).

### 2023-06-27 analysis/openarchaeo_collaboration.qmd
Revise fig-cats-plats
Properly formatted bibliographic references

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo (https://open-archaeo.info/), a curated list of 500+ active open projects in archaeology, augmented where available with data on their activity on GitHub. We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations. Our results show an uneven adoption of open source collaborative practices beyond the basic use of `git` as a version control system and GitHub to host source code. A majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base. More community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories. Preliminary findings from ongoing work also shows how collaborative behaviours vary across different subcommunities. By analyzing interactions among archaeologists using the GitHub platform, including the ways they contribute to collective code bases, we demonstrate how archaeological software engineering is beginning to foster new kinds of collaborative commitments that blend open source protocols, DIY work ethic, conventional forms of scholarly communication, and cultural practices unique to archaeology.
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---

@@

```{r plot-theme}
theme_oarch <- function() theme(
)
```

>>>>

  In this paper, we investigate modes of collaboration in this emerging community of practice using open-archaeo (https://open-archaeo.info/), a curated list of 500+ active open projects in archaeology, augmented where available with data on their activity on GitHub. We apply network analysis to map the networks of collaboration that have emerged among archaeologists using GitHub, and conduct a quantitative exploratory analysis to characterize the nature and intensity of these collaborations. Our results show an uneven adoption of open source collaborative practices beyond the basic use of `git` as a version control system and GitHub to host source code. A majority of repositories have 1–3 contributors, with only a few projects distinguished by an active and diverse developer base. More community-oriented features, such as GitHub's 'issues' and 'comments', are used in only a minority of repositories. Preliminary findings from ongoing work also shows how collaborative behaviours vary across different subcommunities. By analyzing interactions among archaeologists using the GitHub platform, including the ways they contribute to collective code bases, we demonstrate how archaeological software engineering is beginning to foster new kinds of collaborative commitments that blend open source protocols, DIY work ethic, conventional forms of scholarly communication, and cultural practices unique to archaeology.
bibliography: openarchaeo-collaboration.bib
format:
  html:
    page-layout: article
    toc: true
    number-sections: true
    embed-resources: true
---

@@

```{r plot-theme}
theme_oarch <- function() {
  theme_minimal() +
  theme()
}
```
